---
title: "p8105_hw2_hs3653"
author: "Hyder Said"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(janitor)
```

## Problem 1
```{r}
subway_df = 
  read_csv(
    "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    col_types = cols(
      Route8 = "c",
      Route9 = "c",
      Route10 = "c",
      Route11 = "c"
    )
  ) |>
  janitor::clean_names() |>
  select(
    line, station_name, station_latitude, station_longitude,
    starts_with("route"), entry, exit_only, vending,
    entrance_type, ada
  ) |>
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )

glimpse(subway_df)
```

# Rows and columns in the cleaned dataset
```{r}
dim(subway_df)
```

# Distinct stations, identified by both line and station name
```{r}
subway_df |>
  distinct(line, station_name) |>
  nrow()
```

# Distinct ADA-compliant stations
```{r}
subway_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()
```


# Proportion of entrances/exits without vending that allow entry
```{r}
subway_df |>
  filter(vending == "NO") |>
  summarise(proportion_allow_entry = mean(entry))
```

The dataset contains information on 1,868 subway entrances and exits, including station names, lines, station coordinates, routes served, entry availability, vending, entrance type, exit-only status, and ADA compliance. I standardized column names, retained the relevant variables, specified character types for route columns 8-11, and converted entry availability to a logical variable. The cleaned dataset contains 1,868 rows and 20 columns. These data are not yet tidy because routes served are spread across 11 columns.

There are 465 distinct stations, identified by station name and line, of which 84 are ADA compliant. Among station entrances and exits without vending, 37.7% allow entry. After reshaping the route columns into route-number and route-name variables, 60 distinct stations serve the A train, of which 17 are ADA compliant.

## Problem 2

## Problem 3