---
title: "p8105_hw2_hs3653"
author: "Hyder Said"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(janitor)
```

## Problem 1
```{r}
subway_df = 
  read_csv(
    "data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
    col_types = cols(
      Route8 = "c",
      Route9 = "c",
      Route10 = "c",
      Route11 = "c"
    )
  ) |>
  janitor::clean_names() |>
  select(
    line, station_name, station_latitude, station_longitude,
    starts_with("route"), entry, exit_only, vending,
    entrance_type, ada
  ) |>
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )

glimpse(subway_df)
```

# Rows and columns in the cleaned dataset
```{r}
dim(subway_df)
```

# Distinct stations, identified by both line and station name
```{r}
subway_df |>
  distinct(line, station_name) |>
  nrow()
```

# Distinct ADA-compliant stations
```{r}
subway_df |>
  filter(ada == TRUE) |>
  distinct(line, station_name) |>
  nrow()
```


# Proportion of entrances/exits without vending that allow entry
```{r}
subway_df |>
  filter(vending == "NO") |>
  summarise(proportion_allow_entry = mean(entry))
```

The dataset contains information on 1,868 subway entrances and exits, including station names, lines, station coordinates, routes served, entry availability, vending, entrance type, exit-only status, and ADA compliance. I standardized column names, retained the relevant variables, specified character types for route columns 8-11, and converted entry availability to a logical variable. The cleaned dataset contains 1,868 rows and 20 columns. These data are not yet tidy because routes served are spread across 11 columns.

There are 465 distinct stations, identified by station name and line, of which 84 are ADA compliant. Among station entrances and exits without vending, 37.7% allow entry. After reshaping the route columns into route-number and route-name variables, 60 distinct stations serve the A train, of which 17 are ADA compliant.

## Problem 2
```{r}
mr_trash_df = 
  read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    range = cell_cols("A:N")
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    across(
      c(dumpster, weight_tons, volume_cubic_yards,
        plastic_bottles, polystyrene, cigarette_butts,
        glass_bottles, plastic_bags, wrappers, homes_powered),
      as.numeric
    ),
    year = as.integer(year),
    date = as.Date(date),
    sports_balls = as.integer(round(sports_balls)),
    wheel = "Mr. Trash Wheel"
  )

professor_trash_df = 
  read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    range = cell_cols("A:M")
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    across(
      c(dumpster, weight_tons, volume_cubic_yards,
        plastic_bottles, polystyrene, cigarette_butts,
        glass_bottles, plastic_bags, wrappers, homes_powered),
      as.numeric
    ),
    year = as.integer(year),
    date = as.Date(date),
    wheel = "Professor Trash Wheel"
  )

gwynnda_df = 
  read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Gwynnda the Good Wheel of the W",
    range = cell_cols("A:L")
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    across(
      c(dumpster, weight_tons, volume_cubic_yards,
        plastic_bottles, polystyrene, cigarette_butts,
        plastic_bags, wrappers, homes_powered),
      as.numeric
    ),
    year = as.integer(year),
    date = as.Date(date),
    wheel = "Gwynnda"
  )

trash_wheel_df = 
  bind_rows(mr_trash_df, professor_trash_df, gwynnda_df)

trash_wheel_df |>
  count(wheel)

glimpse(trash_wheel_df)
```

Collect totals
```{r}
professor_total_weight = 
  trash_wheel_df |>
  filter(wheel == "Professor Trash Wheel") |>
  summarise(total = sum(weight_tons, na.rm = TRUE)) |>
  pull(total)

gwynnda_june2022_butts = 
  trash_wheel_df |>
  filter(
    wheel == "Gwynnda",
    year == 2022,
    month == "June"
  ) |>
  summarise(total = sum(cigarette_butts, na.rm = TRUE)) |>
  pull(total)

professor_total_weight
gwynnda_june2022_butts
```

The combined dataset contains `r nrow(trash_wheel_df)` dumpster collection observations from Mr. Trash Wheel, Professor Trash Wheel, and Gwynnda. Key variables include collection date, trash weight in tons, volume in cubic yards, and counts of litter such as plastic bottles and cigarette butts. I standardized column names, removed rows without dumpster numbers, harmonized variable types, and rounded sports-ball counts to integers. A wheel identifier distinguishes observations from the three sources. Based on available data, Professor Trash Wheel collected `r round(professor_total_weight, 2)` tons of trash, and Gwynnda collected `r format(gwynnda_june2022_butts, big.mark = ",", scientific = FALSE, trim = TRUE)` cigarette butts in June 2022.

## Problem 3